---
title: "Exercise 1 - Getting started with data science"
author: "Your name here"
output: html_document
---

Assignments for the course focus on practical aspects of the concepts covered in the lectures. Assignments are based on the material covered in James et al. You will start working on the assignment in the lab sessions after the lectures, but may need to finish them after class.  If you do not get through the whole assignment, do not worry!

## 1. Working with a sample dataset

This exercise relates to the `College` data set, which can be found in the file `College.csv` on the website for the main course textbook (James et al 2013) https://www.statlearning.com/resources-first-edition. It contains a number of variables for 777 different universities and colleges in the US.   

The variables are:    
* `Private` : Public/private indicator  
* `Apps` : Number of applications received  
* `Accept` : Number of applicants accepted  
* `Enroll` : Number of new students enrolled  
* `Top10perc` : New students from top 10% of high school class   
* `Top25perc` : New students from top 25% of high school class  
* `F.Undergrad` : Number of full-time undergraduates  
* `P.Undergrad` : Number of part-time undergraduates  
* `Outstate` : Out-of-state tuition  
* `Room.Board` : Room and board costs  
* `Books` : Estimated book costs  
* `Personal` : Estimated personal spending  
* `PhD` : Percent of faculty with Ph.D.'s  
* `Terminal` : Percent of faculty with terminal degree  
* `S.F.Ratio` : Student/faculty ratio  
* `perc.alumni` : Percent of alumni who donate  
* `Expend` : Instructional expenditure per student  
* `Grad.Rate` : Graduation rate  

Before reading the data into R, it can be viewed in Excel or a text editor, if you find that convenient.

### 1.1 Load the data

Use the `read.csv()` function to read the data into `R`. Call the loaded data `college`. Make sure that you have the directory set to the correct location for the data.  You can load this in R directly from the website, using:

```{r}
college <- read.csv("https://www.statlearning.com/s/College.csv")
```

### 1.2 View the data

Look at the data using the `View()` function. You should notice that the first column is just the name of each university. We don't really want R to treat this as data. However, it may be handy to have these names for later. Try the following commands:

```{r}
rownames(college) <- college[, 1] 
View(college)
```

You should see that there is now a `row.names` column with the name of each university recorded. This means that R has given each row a name corresponding to the appropriate university. R will not try to perform calculations on the row names. However, we still need to eliminate the first column in the data where the names are stored. Try

```{r}
college <- college[, -1] 
View(college)
```

Now you should see that the first data column is `Private`. Note that another column labeled `row.names` now appears before the `Private` column. However, this is not a data column but rather the name that `R` is giving to each row.

### 1.3 Try some operations

a.  Use the `summary()` function to produce a numerical summary of the variables in the data set.

```{r}
summary(college)
```

   
b.  Use the `pairs()` function to produce a scatterplot matrix of the first ten columns or variables of the data. Recall that you can reference the first ten columns of a matrix `A` using `A[,1:10]`.

```{r}
pairs(college[, 2:11])

# [1:10] won't work because first column is a character variable (text) 
str(college)

# integer is different from number due to decimal points 

```

Also, bear in mind that instead of pairs, if you wanted to call on those columns or rows, think of it like an x,y axis:
```{r}
college[, 2:11] # columns
college[2:11,] # rows

college[2,2] #row 2, column 1
head(college)

# Keep this in mind when using functions especially like plot() 
```
   
c.  Use the `plot()` function to produce side-by-side boxplots of `Outstate` versus `Private`.

```{r}
plot(as.factor(college$Private), college$Outstate,
     xlab = "Private University", ylab = "Tuition in $")
```


d.  Create a new qualitative variable, called `Elite`, by *binning* the `Top10perc` variable. We are going to divide universities into two groups based on whether or not the proportion of students coming from the top 10% of their high school classes exceeds 50%.

```{r} 
Elite <- rep("No", nrow(college))
#rep() is "replicate elements
Elite[college$Top10perc > 50] <- "Yes"
Elite <- as.factor(Elite)
college <- data.frame(college, Elite)
```

Use the `summary()` function to see how many elite universities there are. Now use the `plot()` function to produce side-by-side boxplots of `Outstate` versus `Elite`.
   
   
```{r}
summary(Elite)

plot(college$Elite, college$Outstate, 
     xlab = "Elite University", ylab = "Tuition in $")
```

Continue exploring the data, and provide a brief summary of what you discover.




### 2. Working with the automotive dataset

This exercise involves the `Auto` data set available as `Auto.csv` from the website for the main course textbook James et al. https://www.statlearning.com/resources-first-edition. Make sure that the missing values have been removed from the data. You should load that dataset as the first step of the exercise. 

```{r}

Auto <- read.csv("https://www.statlearning.com/s/Auto.csv", 
                 header = TRUE, na.strings = "?")

# look at the documentation to understand arguments
# without na.strings, $horsepower would be read in as a character variable
?read.csv

#removing NA's
table(is.na(Auto))
Auto <- na.omit(Auto)
#removes 5 rows
table(is.na(Auto))

dim(Auto)
summary(Auto)
```

(a) Which of the predictors are quantitative, and which are qualitative?

Note: Sometimes when you load a dataset, a qualitative variable might have a numeric value.  For instance, the `origin` variable is qualitative, but has integer values of 1, 2, 3.  From mysterious sources (Googling), we know that this variable is coded `1 = usa; 2 = europe; 3 = japan`.  So we can covert it into a factor, using:

```{r}
Auto$originf <- factor(Auto$origin, labels = c("usa", "europe", "japan"))
# attributing the order of these to 1:3
table(Auto$origin)

table(Auto$originf)

# if you want to view the two in context (for instance, to check proper attribution):
with(Auto, table(originf, origin))

```

**Quantitative: mpg, cylinders, displacement, horsepower, weight, acceleration, year. Qualitative: name, origin, originf**

(b) What is the *range* of each quantitative predictor? You can answer this using the `range()` function.

```{r}
#Pulling together qualitative predictors
qualitative_columns <- which(names(Auto) %in% c("name", "origin", "originf"))
qualitative_columns

# Apply the range function to the columns of Auto data
# that are not qualitative
sapply(Auto[, -qualitative_columns], range)

# sapply: Apply a Function over a List or Vector
# instead of range(Auto)

# CONSIDER:
Auto_quant <- Auto[, -qualitative_columns]
# gives you the range for the entire dataset, not each column
sapply(Auto_quant, range)

```

(c) What is the mean and standard deviation of each quantitative predictor?

```{r}
sapply(Auto[, -qualitative_columns], mean)
sapply(Auto[, -qualitative_columns], sd)
```

(d) Now remove the 10th through 85th observations. What is the range, mean, and standard deviation of each predictor in the subset of the data that remains?

```{r}
sapply(Auto[-seq(10, 85), -qualitative_columns], mean)
sapply(Auto[-seq(10, 85), -qualitative_columns], sd)
```

(e) Using the full data set, investigate the predictors graphically, using scatterplots or other tools of your choice. Create some plots highlighting the relationships among the predictors. Comment on your findings.

```{r}
# Part (e):
# pairs(Auto)

pairs(Auto[, -qualitative_columns])
# you will remember this not working earlier if you tried [,1:10]

# Heavier weight correlates with lower mpg.
with(Auto, plot(mpg, weight))
# More cylinders, less mpg.
with(Auto, plot(mpg, cylinders))
# Cars become more efficient over time.
with(Auto, plot(mpg, year))

# Lets plot some mpg vs. some of our qualitative features: 
# sample just 20 observations
Auto.sample <- Auto[sample(1:nrow(Auto), 20), ]
# order them
Auto.sample <- Auto.sample[order(Auto.sample$mpg), ]
# plot them using a "dotchart"
with(Auto.sample, dotchart(mpg, name, xlab = "mpg"))
# same thing, with() enables you not to have to use $
dotchart(Auto.sample$mpg, Auto.sample$name, xlab = "mpg")

with(Auto, plot(originf, mpg, ylab = "mpg"))
plot(Auto$originf, Auto$mpg, ylab = "mpg") # identical to above

```

(f) Suppose that we wish to predict gas mileage (`mpg`) on the basis of the other variables. Do your plots suggest that any of the other variables might be useful in predicting `mpg`? Justify your answer.


**See descriptions of plots in (e). All of the predictors show some correlation with mpg. The name predictor has too little observations per name though, so using this as a predictor is likely to result in overfitting the data and will not generalize well.**

### Further materials

- [Data Camp R tutorials](https://www.datacamp.com/courses/free-introduction-to-r)
- [Data Camp R Markdown tutorials](https://www.datacamp.com/courses/reporting-with-r-markdown). You can complete the free first chapter. 
