---
title: "Base R and Dplyr"
author: "Ken Benoit & Sarah Jewett"
date: "6/22/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Some Basic Base R

```{r}

sqrt(81)
2^4
log(2.718281828)

# round to 2 decimal places
round(2.718281828, 2)

# round up
ceiling(2.718)
# round down
floor(2.718)

#Combine Values or Character Strings into a Vector or List
a <- c(5, 9, 3, 4, 7)

length(a)
#this gives you the length of the vector

# Can then apply functions to them:
sqrt(c(5, 9, 3, 4, 7))
sqrt(a)

#can also use other functions:
sum(c(5, 9, 3, 4, 7))
#   how many?
length(c(5, 9, 3, 4, 7))

#   other statistics
mean(c(5, 9, 3, 4, 7))
median(c(5, 9, 3, 4, 7))
var(c(5, 9, 3, 4, 7))
sd(c(5, 9, 3, 4, 7))  

# use ? to find out what a function does and how to use it properly...
?sd() # parentheses optional

# IMPORTANT!
# The difference between min/max and which.min/which.max
# which() gives you the indices, NOT the value:

# 9 is the largest number
max(c(5, 9, 3, 4, 7))

# 3 is the smallest number
min(c(5, 9, 3, 4, 7))

# 9 is the second number in the sequence, so the output is 2
which.max(c(5, 9, 3, 4, 7))

# 3 is the third number in the sequence, so the output is 3
which.min(c(5, 9, 3, 4, 7))

# retrieve highest number by previously found index:
a <- c(5, 9, 3, 4, 7)
a[2]
c(5, 9, 3, 4, 7)[2]



```

## dplyr

[dplyr](https://dplyr.tidyverse.org) is part of the tidyverse, and it's great because it often simplifies or neatens code that would otherwise become convoluted in base R. Part of dplyr includes:

- Tibbles: transforms data frames into a more palatable way, particularly for viewing in R Studio/RMarkdown. 
  
- Pipe Operators: pipes object first and then functions and arguments after. For example:

        a <- c(5, 9, 3, 4, 7)

        Base R:  min(a)
        dplyr:   a %>% min()

It helps avoid the confusion that comes with a lot of nesting. Let's install dplyr (uncomment the first line if you haven't installed it yet) and then load it. 

```{r message=FALSE}
# install.packages("tidyverse")
library("dplyr")

# note that you are loading just dplyr from the tidyverse, the install downloads all packages associated with the tidyverse and you could also use library(tidyverse) too
```

Let's go over some examples of how to work with dplyr from the link above. It uses Star Wars data that is part of the package. Let's take a look at the data first:

```{r}
data(starwars)

head(starwars)
tail(starwars)
```

**filter()** - picks cases based on values

```{r}
starwars %>% 
  filter(species == "Droid")

# in base R
starwars[which(starwars$species=="Droid"), ]
```

**select()** - pick variables based on the name

```{r}
starwars %>% 
  select(name, ends_with("color"))

#note that when you run ?select, you will see other arguments besides just "ends_with"
```
**mutate()** adds new variables that are functions of existing variables
```{r}
starwars %>% 
  mutate(name, bmi = mass / ((height / 100)  ^ 2)) %>%
  select(name:mass, bmi)
# try running it without the select() function and compare the results

# in Base R 
starwars$bmi <- starwars$mass/(starwars$height/100)^2

```
**arrange()** changes the ordering of rows. It defaults from smallest to largest, so you can use desc()
```{r}
starwars %>% 
  arrange(desc(mass))

# this works like sort() and rev(sort()) in Base R, however it only gives you the values:
(rev(sort(starwars$mass)))

# order does the same thing and you can have the table not unlike dplyr
starwars[order(starwars$mass),]
starwars[rev(order(starwars$mass)),]
# pay attention to what happens to NA's here compared to using dplyr
```

Combining these functions allows you to really manipulate and get to know your data elegantly. When in doubt, slow down and separate by the pipe operator %>% and break down what it is doing one by one:

```{r}
starwars %>%
  group_by(species) %>%
  summarise(
    n = n(),
    # since you are grouping by species, n = n() tallies up the number of occurrences of the species 
    mass = mean(mass, na.rm = TRUE)
  ) %>%
  filter(
    n > 1,
    # now you are specifying that you want to select only those species that have 2 or more occurences in the data. This drops 16 rows of species.
    mass > 50
    # mass being over 50 drops a further 12
  )

```

## tidyR

[tidyR](https://tidyr.tidyverse.org) is great for tidying your data (hence the name!) It is also part of the tidyverse! 

Don't forget that a table is tidy if:

- Each variable is its own column

- Each observation (or case) its own row

We mainly will focus on reshaping with tidyR, namely, pivoting.

Let's go over a few examples from [this resource](https://tidyr.tidyverse.org/articles/pivot.html) together. It uses a datasetfrom a survey asking people about their religion and annual income. 
```{r message=FALSE}
library(tidyr) 

data(relig_income)
head(relig_income)

```
It is currently in a wide format, so we can make it longer using pivot_longer()

```{r}
relig_long <- relig_income %>% 
  pivot_longer(cols = !religion, names_to = "income", values_to = "count")

relig_long
```
#OR

```{r}
relig_income %>% 
  pivot_longer(cols = colnames(relig_income[2:11]), names_to = "income", values_to = "count")
```
Let's pretend we got the data in long form and want it in wide form:

```{r}
relig_wide <- relig_long %>%
  pivot_wider(names_from = c("income"), values_from = "count")
```

As you can see you basically just change the arguments as well as the function itself.

### Let's move on to Assignment 2. You'll need to do some merging, which is part of base R, but here's a hint: there is also the [join() function](https://dplyr.tidyverse.org/reference/join.html) in dplyr 

